Estimation of flow trajectories in a multi-lines transportation network

### 2023-02-27 latex_files/applied_network_science_article/article_v0.tex
finish v0 of article

As case studies, we test the proposed method on toy examples as well as with the data of the public transportation network of the city of Lausanne, in Switzerland. Toy examples offer some kind of validation, as a transportation flow can be entirely set on toy networks and compared to the solution given by our algorithm. The real case scenario with the data from the Lausanne public transportation network demonstrates that this method is applicable on a large transportation network and can yield pertinent insights about traveler habits.

Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the proposed solution. Section \ref{casestudies} contains case studies of the method on toy examples and on the data from the Lausanne public transportation network. Section \ref{conclusion} concludes the article.

@@

The first and second factorial dimensions of the destination and origin profiles of stops, obtained with CA applied on $\mathbf{N}$, can be found on Figure \ref{afc}. We can see that origin and destination coordinates of each stop are quite similar when displayed on the first two factorial axes. This result can be explained if we understand that, in fact, origin and destination profiles are in fact kinda symmetrical: at the beginning of a line, stops have several choices of destinations, but very few origins ``point" at them, which is exactly reversed for stops at the end of a line. Intermediary stops are generally located near city center, and have a similarly rich origin and destination profiles. Concerning the dimensions, we can see that the first one divided the city stops between west and east, a known dichotomy in the public transportation usage in Lausanne: the east of the city is less well served by public transport and inhabitants often use cars, while the west intensively use public transports. The second dimension is harder to interpret, but seems to highlight the radial structure of the transportation network in Lausanne.

Figure \ref{transfers} maps predicted transfer hubs, obtained by aggregating every transfer counts $z_{ij}$ between lines when they occurs in a spatial cluster of stops. As expected, most transfers occur in the city center, with top 5 largest hubs located at St-François (2,132,597 transfers), Lausanne-Gare (2,035,543 transfers), Bel-Air (1,908,168 transfers), Ours (1,057,282 transfers), and Chauderon (969,789 transfers). When taken individually, four of the top 5 transfer counts is predicted between line $1$ and line $72$ occurring at the train station (Lausanne-Gare). Line 72 is actually a subway line connecting the north and south regions of the city, and is by far the most used line in the network. Line $1$ is also a highly frequented line and operated toward the western region of the city. It is not surprising to see most transfer occurring between these two lines, as they permit to connect the region with an high density of public transportation network users (the western part of Lausanne), to the main artery of the network (line 72).

\begin{figure}[h]
	\includegraphics[width=0.46\textwidth]{fig/afc_in_dim1.png}
	\includegraphics[width=0.46\textwidth]{fig/afc_in_dim2.png} \\
	\includegraphics[width=0.46\textwidth]{fig/afc_out_dim1.png}
	\includegraphics[width=0.46\textwidth]{fig/afc_out_dim2.png}
	\caption{Color representations (positive values in red, negative values in blue) of the first two factorial coordinates of stops, after Correspondence analysis (CA) performed on transportation matrix $\mathbf{N}$. Top row maps stops as ``origin profiles" (``where passengers are going"), bottom row maps stops as ``destination profile" (``where passengers are coming from"). Left column is the first factorial dimension, right column is the second factorial dimension. Ring sizes correspond to the number of embarkments (top) and disembarkment (bottom).}
	\label{afc}
\end{figure}

@@

\section{Conclusion}
\label{conclusion}

\newpage

>>>>

As case studies, we test the proposed method on toy examples as well as with the data of the public transportation network\footnote{\url{https://www.t-l.ch/}} of the city of Lausanne, in Switzerland. Toy examples offer some kind of validation, as a transportation flow can be entirely set on toy networks and compared to the solution given by our algorithm. The real case scenario with the data from the Lausanne public transportation network demonstrates that this method is applicable on a large transportation network and can yield pertinent insights about traveler habits.

Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the proposed solution. Section \ref{casestudies} contains case studies of the method on toy examples and on the data from the Lausanne public transportation network. Section \ref{conclusion} concludes the article. Data, codes and extensive results can be found in the GitHub repository of the article\footnote{\url{https://github.com/sliunil/tl_study}}.

@@

The first and second factorial dimensions of the destination and origin profiles of stops, obtained with CA applied on $\mathbf{N}$, can be found on Figure \ref{afc}. We can see that origin and destination coordinates of each stop are quite similar when displayed on the first two factorial axes. This result can be explained if we understand that, in fact, origin and destination profiles are in fact, to some extent, symmetrical: at the beginning of a line, stops have several choices of destinations, but very few origins ``point" at them, which is exactly the reverse for stops at the end of a line. Intermediary stops are generally located near the city center, and have a similarly rich origin and destination profiles. Concerning the dimensions, we can see that the first component divides the city stops between west and east, a known dichotomy in the public transportation usage in Lausanne: the east of the city is less well served by public transport and inhabitants often use cars, while the west intensively uses public transports. The second component is harder to interpret, but seems to highlight the radial structure of the transportation network in Lausanne.

Figure \ref{transfers} maps predicted transfer hubs, obtained by aggregating every transfer counts $z_{ij}$ between lines when they occurs between stops located in the same spatial cluster (same ``superstop" name in the dataset). As expected, most transfers occur in the city center, with the top 5 largest hubs located at St-François (2,132,597 transfers), Lausanne-Gare (2,035,543 transfers), Bel-Air (1,908,168 transfers), Ours (1,057,282 transfers), and Chauderon (969,789 transfers). When taken individually, four of the top 5 transfer counts are predicted between line $1$ and line $72$ occurring at the train station (Lausanne-Gare). Line 72 is actually a subway line connecting the north and south regions of the city, and is by far the most used line in the network. Line $1$ is also a highly frequented line and operated toward the western region of the city. It is not surprising to see most transfer occurring between these two lines, as they permit to connect the region with an high density of public transportation network users (the western part of Lausanne), to the main artery of the network (line 72).

\begin{figure}[h]
	\includegraphics[width=0.46\textwidth]{fig/afc_in_dim1.png}
	\includegraphics[width=0.46\textwidth]{fig/afc_in_dim2.png} \\
	\includegraphics[width=0.46\textwidth]{fig/afc_out_dim1.png}
	\includegraphics[width=0.46\textwidth]{fig/afc_out_dim2.png}
	\caption{Color representations (positive values in red, negative values in blue) of the first two factorial coordinates of stops, obtained from Correspondence analysis (CA) performed on the transportation matrix $\mathbf{N}$. Top row maps stops as ``origin profiles" (``where passengers are going"), bottom row maps stops as ``destination profile" (``where passengers are coming from"). Left column depicts the first factorial dimension, right column the second factorial dimension. Ring sizes correspond to the number of embarkments (top) and disembarkments (bottom).}
	\label{afc}
\end{figure}

@@

\section{Conclusion}
\label{conclusion}
At first glance, the task of estimating the origin-destination trip counts of public transportation networks on the sole basis of embarkments and disembarkments stop counts might appear as too ambitious. The project is indeed challenging, but we hope to have convinced the reader that meaningful estimates can be obtained by applying a succession of carefully chosen, principled yet flexible steps. Some of the ingredients (maximum entropy, iterative fitting) are familiar in transportation theory, and some others (shrinkage of priors, minimal amount of entering and leaving the network) seem original. Possible theoretical and algorithmic improvements are still under investigation.
Due to length constraints, the exploitation and interpretation of the estimated flow has been kept to a minimum. Yet, many issues, such as disaggregating data and regimes (week days, week-ends or holidays; time in the day), assessing the centrality of stops and junctions, or characterizing the imbalance between uphill and downhill flows (Lausanne is known for its slopes) are of primary interest for urban planning and transportation geography, and will be developed in a forthcoming work. Also, visualizing the numerous and various quantities of interest, on a spatial map or else, constitutes a challenge in itself, requiring a particular blend of creativity and rigor.
